\documentclass{article}
\usepackage{graphicx}
\usepackage{natbib}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{booktabs}
\usepackage{array}
\usepackage[margin=1in]{geometry}
\usepackage{hyperref}
\usepackage{xcolor}

\definecolor{notecolor}{rgb}{0.5,0.5,0.5}
\newcommand{\draftnote}[1]{\textit{\textcolor{notecolor}{[#1]}}}

\title{Aligning Incentives in the NBA Draft:\\
       A Multi-Agent Simulation of Mechanism Design}
\author{Grant Valentine \and Kartik Dhinakaran}
\date{May 2026}

\begin{document}

\maketitle
\tableofcontents
\newpage

% ============================================================
%  Section 0: Preface
% ============================================================

\section*{Preface}
\addcontentsline{toc}{section}{Preface}

This project was completed jointly by Grant Valentine and Kartik Dhinakaran.
Both authors read through the full literature.
Kartik took primary responsibility for synthesizing the literature review,
with a focus on the empirical and theoretical literature on tanking incentives
\citep{taylor2002, schmidt2024, lenten2016}.
Grant took primary responsibility for the papers on alternative draft
mechanisms \citep{banchio2020, kazachkov2020, highley2026}.
Grant wrote the formal model.
Kartik wrote the introduction and background section.
Both authors contributed to the simulation code: Grant designed the overall
structure and experimental framework, and Kartik implemented the
mechanism-specific logic for each draft system under study.
Writing and analysis were reviewed by both authors throughout.

No external collaborators contributed to this submission beyond high-level
discussion with other students in the course.

% ============================================================
%  Section 1: Research Question & Core Framing
% ============================================================

\section{Research Question \& Core Framing}

Deliberately losing games to improve future draft position, colloquially known
as tanking, has been a persistent problem in professional sports league design
for decades.
The incentive is not subtle: in 1983, the Houston Rockets deliberately
underperformed late in the season to secure the first overall pick in the
following year's draft, ultimately selecting future Hall-of-Famer Hakeem
Olajuwon.
The NBA responded by restructuring the draft lottery, and has done so again in
1990, 2019, and 2026.
Each reform has failed to resolve the underlying problem.
Tanking remains one of the most maligned phenomena in professional basketball,
criticized by fans, analysts, and the league office alike.

This paper asks: to what degree can differing mechanism designs of the NBA
draft influence the incentive for teams to tank while maintaining league-wide
competitive balance?

The central tension is straightforward.
The NBA draft lottery uses reverse-order weighting to promote competitive
balance, a ``rubber band'' effect designed to steer top incoming talent toward
the weakest franchises.
But this structure creates a well-documented adverse incentive: teams near the
bottom of the standings may find it individually rational to lose games
deliberately in order to improve their lottery odds, particularly when the
incoming draft class contains players with asymmetric superstar potential.
The worse a team finishes, the better its expected draft position.
Winning, in this context, can be a mistake.

From an algorithmic game theory perspective, tanking is structurally a
prisoner's dilemma.
A single team that tanks may improve its own long-run competitive outlook, but
when enough teams tank simultaneously, the benefit to any individual team
washes out while the overall quality of league play degrades.
The end-of-season landscape in recent NBA seasons has become increasingly
bifurcated: teams are either competing for playoff seeding or deliberately
losing to improve draft position, with little in between.
The question is whether any mechanism design can resolve this tension
structurally, or whether the impossibility result of \citet{banchio2020}
forecloses the possibility of a clean solution, leaving any viable mechanism
as a tradeoff rather than a fix.

We examine five mechanisms against this tension: the current NBA lottery
(post-2019 reform), the bilevel ranking system proposed by
\citet{kazachkov2020}, the Carry-Over Lottery Allocation (COLA) framework of
\citet{highley2026}, a weighted-loss mechanism that applies a time-decaying
score to each game result, and a proposed NBA redesign in which the three
worst teams receive \emph{fewer} lottery balls than the teams just above them.
We evaluate each through a combination of formal analysis and a multi-agent
simulation in which teams are represented by agents that develop strategies
endogenously, rather than following hardcoded rules.
This approach lets us test how each mechanism performs against adaptive,
strategically sophisticated agents rather than fixed behavioral assumptions.

% ============================================================
%  Section 2: Introduction & Background
% ============================================================

\section{Introduction \& Background}
\draftnote{Primary author: Kartik. Draft due May 13.}

The NBA is a league consisting of 30 teams, 16 of which make the playoffs each
year.
The remaining 14 non-playoff teams enter a weighted lottery of ping pong balls,
with the composition of balls equivalent to the allotted odds each team has to
pick in the top 4 slots.
From there, slots 5 through 14 are allocated by reverse regular-season record;
slots 15 through 30 occur in the same manner for playoff teams.
The worst pick the team with the worst record could receive is \#5.

This structure has evolved and been re-weighted on multiple occasions by the
NBA:
\begin{itemize}
    \item \textbf{1985:} All 7 non-playoff teams had identical odds at the
    \#1 pick. There is a conspiracy theory surrounding this first lottery
    called the ``frozen envelope'' theory---that Commissioner David Stern, a
    New Yorker, had frozen the NY Knicks envelope so that he would recognize
    it and ensure they selected future Hall of Famer Patrick Ewing.
    \item \textbf{1990:} Weighted the odds by reverse standings for the first
    time; the worst team had a 16.7\% chance at the \#1 pick.
    \item \textbf{1994:} Tweaked so that the worst team has a 25\% chance at
    the \#1 pick.
    \item \textbf{2019:} Top 3 teams' odds at the \#1 pick were reduced to a
    flat 14\%, which had the unintended consequence of making slots 4--7 more
    valuable in relative terms and promoting tanking in a different manner.
\end{itemize}

Tanking is a phenomenon that \citet{taylor2002} and later \citet{schmidt2024}
both quantify and confirm.
Their analyses use regression to show the explanatory power of playoff
elimination on win rates---teams eliminated from playoff contention win at
significantly lower rates than otherwise expected.
Sam Hinkie, GM of the 2013--2016 Philadelphia 76ers, famously coined the
phrase ``The Process'' as his strategy of three consecutive seasons of
stockpiling draft picks and making roster decisions designed to improve draft
position.
He was eventually pushed out by the NBA and ownership due to how unpopular
his strategy was once it became transparent.

The flattening of odds at the top of the curve produced a notable shift in
incentive structure: whereas previously tanking was most valuable for the
handful of worst teams, the post-2019 regime distributes excess value more
evenly across the bottom tier.
The race to the absolute bottom became a race to the bottom \emph{tier}.

\subsection{How the NBA Draft Works}

Each summer, the NBA holds a draft in which teams select incoming players,
primarily from college and international leagues.
The draft is structured to redistribute talent toward weaker franchises: teams
that performed poorly during the regular season receive earlier picks, which
are more likely to be high-impact players.

The allocation of picks is divided into two tiers.
The 16 playoff teams receive picks 17 through 30 in order of their playoff
finish; the 14 non-playoff teams compete for picks 1 through 14 via a lottery
weighted by reverse regular-season record.
Under the current system (since the 2019 reform), the lottery determines only
picks 1 through 4 by weighted draw; picks 5 through 14 are awarded in
reverse-standings order.
The three teams with the worst records each hold a 14.0\% chance at the
first overall pick, followed by the 4th-worst at 12.5\%, declining steeply
to 1.0\% for the team with the 11th-worst record, 0.5\% for the 12th, 0.5\%
for the 13th, and 0.0\% for the 14th.

The lottery has been redesigned several times since its introduction in 1985.
In its original form, all non-playoff teams had equal odds at the first
overall pick---a design that notoriously invited abuse and spawned the
``frozen envelope'' conspiracy theory surrounding the Knicks' selection of
Patrick Ewing.
The 1990 reform introduced weighted odds based on reverse standings, giving
the worst team a 16.7\% chance; the 1994 adjustment raised that figure to
25\%, making the bottom pick more valuable still.
The 2019 reform reversed course: it \emph{flattened} the top of the
distribution by equalizing the top-three teams' odds at 14\% each, hoping
to reduce the absolute value of being the worst team in the league.
As we discuss next, this reform displaced the incentive rather than
eliminating it.

\subsection{Evidence of Tanking}

The empirical literature establishes that tanking is not merely a theoretical
concern but a behavioral reality.
\citet{taylor2002} exploit two lottery-regime changes---the equal-odds lottery
introduced in 1984--85 and the weighted lottery in 1990---to estimate the
causal effect of playoff elimination on team win rates.
Controlling for team quality, eliminated teams were 2.5 times as likely to
lose as predicted in 1983--84, 1.3 times in the equal-odds 1984--85 season,
and 2.2 times again after the introduction of weighted odds.
The effect tracks the incentive structure precisely: it rose when the lottery
rewarded poor records, fell when all teams had equal odds, and rose again when
weighting was restored.

\citet{schmidt2024} extends this analysis to the 2019 reform using a portfolio
theory framework that measures deviation from optimal player deployment.
He finds that lottery-contending teams significantly underdeployed their best
lineups in 2017--18 under the high-concentration pre-reform odds, and that
this effect largely disappeared after the 2019 flattening.
Notably, the tanking effect is substantially larger for road games, consistent
with teams being less willing to tank visibly in front of home fans.

Together, these papers confirm that tanking responds to incentives in the
manner that theory predicts, and that it is quantifiable, not anecdotal.

\subsection{Why Existing Mechanisms Keep Failing}

The 2019 reform illustrates a persistent problem with parameter-tuning
approaches: flattening the top of the odds distribution did not eliminate
tanking; it relocated it.
By equalizing odds among the three worst teams, the reform made the 4th
through 7th worst records more valuable in relative terms.
Teams gained more from finishing in this ``sweet spot'' than from competing
for a playoff berth, producing a second-tier tanking race that replaced the
original one.

\citet{banchio2020} provide a formal explanation for why this pattern recurs.
Their central result (Theorem 1) is an impossibility theorem: any mechanism
that (a) allocates picks based solely on end-of-season standings, and (b)
does not treat all non-playoff teams identically, will give some team an
incentive to lose in some possible season history.
The proof is by contradiction: if the mechanism is not flat and respects
standings, there always exists a position at which losing improves expected
pick quality by more than it costs in playoff probability.
Crucially, this impossibility holds regardless of how the weighting is
calibrated---it is a structural property of end-of-season ranking systems,
not a consequence of any specific parameter choice.

The implications are important for mechanism design.
No tweak to the current system can resolve the tanking problem without
either abandoning the standings-based ranking structure entirely or treating
all lottery teams equally (which would destroy redistributive efficiency,
since top picks would go to teams at random rather than to the weakest
franchises).
The reform proposals we study---bilevel ranking, COLA, weighted loss, and the
NBA's 3-2-1 proposal---each address this constraint in a different way, which
is what makes comparing them informative.

% ============================================================
%  Section 3: Formal Model
% ============================================================

\section{Formal Model}

This section defines the formal quantities used throughout the paper.

\subsection{Agents and Utilities}

Teams $i = 1, \ldots, n$ are the agents.
Each team $i$ has a true skill level $\alpha_i \in [0,1]$, a nonlinear
function of roster quality that is treated as exogenous.

Each team's utility comes from one of two mutually exclusive outcomes.
If a team qualifies for the playoffs, it receives a playoff value $V$,
representing the revenue, prestige, and long-run competitive value of
postseason play.
If a team does not qualify, it enters the draft lottery and receives the
value of its draft pick, $D(k)$, where $k$ is the pick number and
$D(\cdot)$ is strictly decreasing in $k$.
Making the playoffs precludes lottery entry; the two payoffs are exclusive.

The utility function for team $i$ is therefore:
\[
    U_i =
    \begin{cases}
        V & \text{if team $i$ makes the playoffs,} \\
        \mathbb{E}[D(k_i)] & \text{otherwise,}
    \end{cases}
\]
where $k_i$ is the draft pick drawn by the mechanism from the lottery over
all non-playoff teams.
Writing $\pi_i$ for team $i$'s ex-ante probability of making the playoffs,
the expected utility under any strategy profile is
\[
    \mathbb{E}[U_i] = \pi_i \cdot V
                    + (1 - \pi_i) \cdot \mathbb{E}[D(k_i) \mid \text{lottery}],
\]
where the expectation in the second term is over the lottery randomness
given the team's expected standing among non-playoff teams.

\subsection{When Is Tanking Individually Rational?}

A team has an incentive to tank when the marginal gain in expected draft pick
value from losing exceeds the marginal loss in playoff probability:
\[
    \frac{\partial}{\partial e_i}\,\mathbb{E}[D(k)] \cdot p_i < 0
    \quad \Longleftrightarrow \quad
    \text{tanking is individually rational for team } i.
\]
This incentive is concentrated around the playoff cutoff, where the cliff in
$V$ is steepest.
Teams safely in the playoffs face no tanking incentive; teams far from playoff
contention face a large incentive; teams near the bubble face the sharpest
tradeoff.
We formally characterize this condition under each lottery odds structure in
Section~\ref{sec:conditions}.

\subsection{Mechanisms Under Study}

Table~\ref{tab:mechanisms} summarizes the five mechanisms evaluated in this
paper.

\begin{table}[h]
\centering
\renewcommand{\arraystretch}{1.4}
\begin{tabular}{p{3.8cm} p{3.8cm} p{5.4cm}}
\toprule
\textbf{Mechanism} & \textbf{Lottery Basis} & \textbf{Key Property} \\
\midrule
Current NBA lottery (2019) &
End-of-season standings, flattened weights &
Tanking incentive exists by \citet{banchio2020}; top-3 each at 14\% \\
Bilevel ranking \citep{kazachkov2020} &
Standings at breakpoint $\sim$5/6--7/8 through season &
Post-breakpoint tanking yields no benefit; 57--72\% reduction \\
COLA \citep{highley2026} &
Multi-year ticket accumulation; equal per-season allocation &
Fully incentive-compatible in-season; tickets carry over \\
Weighted loss &
Decaying per-game loss score; early losses count more &
Suppresses late-season tanking continuously, no hard cutoff \\
NBA ``3-2-1'' proposal &
Tier-based: bottom 3 get 2 balls, next 11 get 3 balls &
Inverts extreme tanking incentive; Tier~A floor at pick~12 \\
\bottomrule
\end{tabular}
\caption{Five mechanisms under study, lottery basis, and key properties.}
\label{tab:mechanisms}
\end{table}

% ============================================================
%  Section 4: Literature Review
% ============================================================

\section{Literature Review}

\draftnote{Primary synthesis author: Kartik. Individual paper notes provided
below. The cross-paper synthesis---comparing, contrasting, and critiquing
across papers---is where the AGT contribution lives and should be written
after all individual sections are drafted.}

\subsection{\citet{banchio2020}: ``A No-Tanking Draft Allocation Policy''}

\textbf{In one sentence:} This paper identifies that since any end-of-year
ranking system will create perverse incentives, the NBA should only account
for games up to the point at which a team first has an incentive to tank.

\subsubsection*{Setting \& Model}
\begin{itemize}
    \item Teams are the agents, each choosing an effort level that affects
    their probability of winning individual games and, ultimately, their
    end-of-season rank.
    \item The season is modeled as a sequence of pairwise games; team effort
    translates to win probability through a parameter $\alpha$.
    \item The league is modeled as a central planner whose objective is to
    maximize the probability that the first draft pick is assigned to the
    worst team, subject to the constraint that no team has an incentive to
    tank at any point in the season.
    \item The proposed mechanism updates lottery probabilities dynamically
    after each game, setting each team's probability to match its current
    likelihood of finishing last given the season history so far.
    \item This adjustment process is terminated before continuing it would
    violate incentive-compatibility for any team---meaning the cutoff is
    endogenous to the season's trajectory, not fixed in advance.
\end{itemize}

\subsubsection*{Main Results}
\begin{itemize}
    \item \textbf{Impossibility (Theorem 1):} Any redistributive policy that
    relies on final standings alone to allocate draft picks will give some team
    an incentive to lose in some possible season history. Proved by
    contradiction. This is the key negative result motivating the whole
    literature.
    \item \textbf{Positive result:} Their dynamic mechanism sidesteps the
    impossibility by locking in lottery probabilities mid-season, before
    tanking becomes rational. It is approximately optimal among all
    incentive-compatible rules that operate on the same restricted set of games.
    \item The mechanism does not guarantee the best player goes to the worst
    team, but approximates it far better than a flat uniform lottery would.
\end{itemize}

\subsubsection*{Proof Techniques / Methodology}
\begin{itemize}
    \item Theorem 1: proof by contradiction.
    \item Simulation of a single season at a time to identify the cutoff point.
    \item Retroactively applied to the 1987 and 1989 NBA seasons---smart
    because flat lottery odds in those seasons means the data is not
    contaminated by teams already behaving strategically.
\end{itemize}

\subsubsection*{Relation to Our Project}
\begin{itemize}
    \item Their model setup is very robust; we implement a simplified version.
    \item We adopt their framework for a value from the draft $D$ and a value
    from making the playoffs $V$.
    \item We do not use their $\alpha$ parameter for effort; instead we use an
    effort floor model in which teams cannot perfectly throw a game even at
    minimum effort.
\end{itemize}

\subsubsection*{Connections to Other Papers}
\begin{itemize}
    \item The impossibility result (Theorem 1) is the direct theoretical
    foundation for both \citet{kazachkov2020} and \citet{highley2026}---both
    papers respond by designing mechanisms that do not rely on final standings.
    \item Banchio \& Munro explicitly discuss and distinguish their proposal
    from Gold (2010) and \citet{lenten2016}; they argue neither fully resolves
    the incentive problem.
    \item Agrees with \citet{taylor2002} that tanking is a real, documented
    phenomenon, not just a theoretical concern.
\end{itemize}

\subsubsection*{Critiques / Open Questions}
\begin{itemize}
    \item To calculate when draft odds should be locked in, the league must
    know both $D$ and $V$ for each team---in practice these are unobservable
    and likely heterogeneous across franchises.
    \item The game at which draft order is set is not transparent or
    predictable in advance, which creates fan engagement and trust problems.
    \item Ignores game outcomes from the latter part of the season entirely,
    which may reduce competitive effort even absent intentional tanking---teams
    whose position is locked in may rest healthy players to avoid injury.
    \item No incentive for teams to continue playing hard after the cutoff
    point is reached.
\end{itemize}

\subsection{\citet{kazachkov2020}: ``On Tanking and Competitive Balance''}

\textbf{In one sentence:} This paper introduces a bilevel ranking mechanism
that determines non-playoff draft order by standings at a mid-season
breakpoint rather than end of season, proving it both reduces tanking and can
improve competitive balance relative to the current NBA system.

\subsubsection*{Setting \& Model}
\begin{itemize}
    \item A stylized sports league with $n$ teams, each with a true (latent)
    quality level the league cannot directly observe.
    \item Teams choose whether to exert full effort or tank in each game;
    observed standings are a signal of true quality that is distorted when
    teams tank.
    \item Two belief models: an optimistic model (a tanking team always
    benefits) and a conservative model (gains depend on other teams'
    behavior)---the dominant strategy result holds under both.
    \item Two evaluation metrics: tanking rate and competitive balance,
    measured by Kendall tau distance between the final draft ordering and the
    true quality ranking.
    \item The bilevel mechanism ranks non-playoff teams by standings at a
    preset mid-season breakpoint; once the breakpoint passes, draft position
    is locked in.
\end{itemize}

\subsubsection*{Main Results}
\begin{itemize}
    \item Under the current NBA system, tanking is a dominant strategy for
    any team effectively eliminated from playoff contention, proved under both
    belief models.
    \item Under the bilevel mechanism, no team has an incentive to tank in any
    game after the breakpoint has passed.
    \item Simulation results show the bilevel ranking reduces tanked games by
    57--72\% compared to the current NBA system, depending on breakpoint
    placement.
    \item Counterintuitively, the bilevel mechanism can also improve
    competitive balance relative to the current system: late-season tanking
    distorts standings; eliminating it produces a draft ordering that better
    reflects which teams genuinely need help.
    \item The optimal breakpoint falls between 5/6 and 7/8 of the way through
    the season.
\end{itemize}

\subsubsection*{Proof Techniques / Methodology}
\begin{itemize}
    \item Formal game-theoretic analysis to establish dominance results under
    both belief models.
    \item Closed-form characterization of team tanking strategies under
    bilevel ranking.
    \item Monte Carlo simulation of full league seasons with teams playing
    fixed strategies.
    \item Kendall tau distance used throughout as the competitive balance
    metric---borrowed directly for our simulation.
\end{itemize}

\subsubsection*{Relation to Our Project}
\begin{itemize}
    \item The bilevel ranking is Condition D in our experimental design.
    \item We borrow their Kendall tau distance metric as our primary measure
    of competitive balance.
    \item Our simulation tests whether their 57--72\% tanking reduction holds
    when agents are adaptive rather than rule-based---this is our main
    extension of their work.
\end{itemize}

\subsubsection*{Connections to Other Papers}
\begin{itemize}
    \item Accepts the \citet{banchio2020} impossibility result as its starting
    point; the bilevel mechanism circumvents it by not relying on final
    standings.
    \item Outperforms the \citet{lenten2016} elimination-based ordering in
    direct simulations.
    \item \citet{highley2026} takes a more structural approach to the same
    problem; bilevel ranking keeps the single-season structure while COLA
    introduces multi-year carryover.
\end{itemize}

\subsubsection*{Critiques / Open Questions}
\begin{itemize}
    \item The breakpoint must be chosen in advance; the paper does not provide
    a principled method for determining the optimal breakpoint, which likely
    varies by season.
    \item Teams could tank strategically before the breakpoint to position
    themselves favorably---the mechanism eliminates post-breakpoint incentives
    but not pre-breakpoint ones.
    \item The model treats team quality as fixed; mid-season trades complicate
    the breakpoint calculation in practice.
    \item Their simulations use rule-based agents with fixed strategies---our
    LLM-based simulation directly addresses this limitation.
\end{itemize}

\subsection{\citet{highley2026}: ``Carry-Over Lottery Allocation''}

\textbf{In one sentence:} COLA achieves full incentive compatibility by giving
every non-playoff team the same number of lottery tickets each season
regardless of record, while still favoring persistently weak teams through
ticket carryover across seasons.

\subsubsection*{Setting \& Model}
\begin{itemize}
    \item Multi-season model; teams accumulate a stockpile of lottery tickets
    over time.
    \item Each season, every non-playoff team receives the same fixed number
    of new tickets regardless of win-loss record---the core design choice that
    eliminates the in-season tanking incentive.
    \item Tickets carry over into future lotteries; playoff success and winning
    a top pick reduce a team's stockpile.
    \item Classic COLA is the primary proposal; Diet COLA is a simplified
    variant used for formal proofs.
    \item Special case: in years with an exceptionally strong draft class,
    some playoff teams might prefer lottery access over their playoff spot.
    The paper handles this via a Bayesian Truth Serum-based media survey to
    identify such years, expanding lottery eligibility to maintain
    incentive-compatibility.
\end{itemize}

\subsubsection*{Main Results}
\begin{itemize}
    \item Classic COLA eliminates the in-season tanking incentive entirely:
    ticket allocation does not depend on record.
    \item The carryover structure preserves the competitive balance function
    of the draft---teams that repeatedly miss the playoffs accumulate more
    tickets and are more likely to land top picks over time.
    \item Diet COLA is formally proved incentive-compatible under the
    assumption that reaching the playoffs is always preferred to lottery
    access.
    \item The Bayesian Truth Serum mechanism for identifying strong draft
    classes is a notable methodological contribution in its own right.
    \item Practical implementation challenges addressed: transitioning from
    the current system, handling traded picks (excluded from lottery
    eligibility), managing varying draft class strength.
\end{itemize}

\subsubsection*{Proof Techniques / Methodology}
\begin{itemize}
    \item Formal proof of incentive-compatibility for Diet COLA under stated
    preference assumptions.
    \item Mechanism design analysis for the strong draft class extension using
    Bayesian Truth Serum.
    \item Qualitative analysis of implementation challenges and practical
    feasibility.
\end{itemize}

\subsubsection*{Relation to Our Project}
\begin{itemize}
    \item COLA is Condition E in our experimental design.
    \item Our primary contribution relative to this paper is testing whether
    the IC guarantee holds when agents are adaptive rather than idealized
    rational agents.
    \item The multi-year ticket accumulation mechanic presents a design
    challenge for our single-season simulation; we run multi-season simulations
    to let ticket stockpiles develop endogenously.
\end{itemize}

\subsubsection*{Connections to Other Papers}
\begin{itemize}
    \item Accepts the \citet{banchio2020} impossibility result as foundational
    motivation; COLA's response is the most structurally radical---it abandons
    standings as the basis for ticket allocation entirely.
    \item Makes a stronger IC guarantee than \citet{kazachkov2020} (full IC
    rather than partial reduction) but requires multi-year infrastructure that
    bilevel ranking does not.
    \item Both COLA and bilevel ranking favor simpler, more transparent rules
    over the dynamic adjustment approach of \citet{banchio2020}.
\end{itemize}

\subsubsection*{Critiques / Open Questions}
\begin{itemize}
    \item Multi-year ticket carryover introduces significant administrative
    complexity for teams, fans, and analysts.
    \item The Bayesian Truth Serum mechanism may itself be manipulable if
    media respondents have team affiliations or financial incentives.
    \item The IC proof depends on the assumption that all teams always prefer
    playoff access to lottery access outside of specially-handled strong-class
    years; if this assumption fails more broadly, the IC guarantee weakens.
    \item Excluding traded picks from lottery eligibility creates a two-tier
    system that could distort the trade market in unmodeled ways.
\end{itemize}

\subsection{\citet{taylor2002}: ``Losing to Win''}

\draftnote{Primary author: Kartik. Use narrowly to establish that tanking
incentives produce real behavior. Do not let it drive the paper's framing.
This is AGT, not sports economics.}

\textbf{In one sentence:} An early empirical paper that quantified tanking
by exploiting two lottery-regime changes to show that eliminated teams' win
rates correlate strongly with the incentive structure of the active lottery.

\subsubsection*{Setting \& Model}
\begin{itemize}
    \item Three NBA seasons: 1983--84 (pre-lottery), 1984--85 (equal-odds
    lottery), and 1990 (weighted lottery). Each observation is a single
    regular-season game.
    \item Response variable: binary win/loss.
    \item Key question: controlling for team quality, do playoff-eliminated
    teams lose more, and does the effect differ significantly across lottery
    regimes?
\end{itemize}

\subsubsection*{Main Results}
\begin{itemize}
    \item 1983--84: eliminated teams 2.5$\times$ as likely to lose vs.
    expected from team quality.
    \item 1984--85: flattened odds reduce the effect to 1.3$\times$.
    \item 1990: eliminated teams 2.2$\times$ as likely to lose.
    \item Lottery structure significantly impacts the propensity of eliminated
    teams to lose.
\end{itemize}

\subsubsection*{Proof Techniques / Methodology}
\begin{itemize}
    \item Empirical logit regression; econometric rather than
    mechanism-design or game-theoretic.
\end{itemize}

\subsubsection*{Relation to Our Project}
\begin{itemize}
    \item Provides compelling evidence that tanking is a real effect driven
    by incentives, not just a theoretical concern. Seminal paper in the
    literature.
\end{itemize}

\subsubsection*{Critiques / Open Questions}
\begin{itemize}
    \item Limited sample size of 3 seasons; particularly constrained for the
    equal-odds regime (1 season only).
    \item Game outcomes do not necessarily capture fan utility---fans may value
    effort and star-player appearances regardless of wins.
    \item Likely underestimates modern tanking given how analytically
    sophisticated front offices have become in the 24 years since publication.
\end{itemize}

\subsection{\citet{schmidt2024}: ``On the Incentive Structure of Tournaments''}

\draftnote{Primary author: Kartik. Same caveat as Taylor \& Trogdon---
empirical background only.}

\textbf{In one sentence:} A successor to \citet{taylor2002} that uses modern
portfolio theory to show that the NBA's 2019 flattening of top-pick odds led
to a measurable decline in asset deployment quality among lottery-contending
teams, with effects disappearing after the reform.

\subsubsection*{Setting \& Model}
\begin{itemize}
    \item Two NBA seasons: 2017--18 (pre-reform) and 2018--19 (post-reform).
    The split is the flattening of \#1 pick odds from 25\%/19.9\%/15.6\% to
    14\% flat for the three worst teams.
    \item Frames lineup decisions as an asset allocation problem; measures
    the ``return differential'' between realized and theoretically optimal
    player deployment for lottery-contending teams.
\end{itemize}

\subsubsection*{Main Results}
\begin{itemize}
    \item 2017--18: measurably more likely that lottery-contending teams would
    underperform their optimal lineup allocation.
    \item 2018--19: effect disappears after odds are flattened.
    \item Interesting home/away split: the effect is much larger on the road,
    suggesting teams are reluctant to tank visibly in front of home fans.
\end{itemize}

\subsubsection*{Proof Techniques / Methodology}
\begin{itemize}
    \item Modern portfolio theory to construct an efficient frontier of player
    allocations per game; regression on the deviation between realized and
    theoretical return.
\end{itemize}

\subsubsection*{Relation to Our Project}
\begin{itemize}
    \item Provides additional evidence that tanking is incentive-driven.
    The 2019 reform is the most recent real-world data point directly
    relevant to our simulation's current-NBA baseline.
\end{itemize}

\subsubsection*{Critiques / Open Questions}
\begin{itemize}
    \item Limited sample size of 2 seasons.
    \item Neither this paper nor \citet{taylor2002} captures the effect of
    incoming draft class strength---teams may tank more in generational draft
    years, which is left unmodeled.
\end{itemize}

\subsection{\citet{lenten2016}: ``Mitigation of Perverse Incentives in
Professional Sports Leagues''}

\draftnote{Primary author: Kartik. Broader comparative context across sports;
useful for framing our mechanisms within a wider design space.}

\textbf{In one sentence:} Proposes assigning draft picks by order of playoff
elimination, arguing this locks in draft position before teams have incentive
to lose.

\subsubsection*{Setting \& Model}
\begin{itemize}
    \item Frames reverse-order drafts as an industrial/organizational problem.
    \item Claims: once a team is eliminated, its draft position is locked in,
    removing post-elimination incentive to lose.
    \item MLB and NBA seasons from 2005--2013; Monte Carlo simulations to
    estimate elimination timing.
\end{itemize}

\subsubsection*{Main Results}
\begin{itemize}
    \item Controlling for other factors, teams play meaningfully better in
    the post-elimination window than the pre-elimination window, holding
    across both NBA and MLB.
\end{itemize}

\subsubsection*{Proof Techniques / Methodology}
\begin{itemize}
    \item Monte Carlo simulation of remaining-schedule outcomes to estimate
    each team's elimination date.
    \item Regression of game outcomes on pre- vs.\ post-elimination behavior.
\end{itemize}

\subsubsection*{Relation to Our Project}
\begin{itemize}
    \item Provides a unique mechanism idea empirically validated within the
    paper; findings reinforce our project's core motivation that incentive
    changes produce behavioral changes.
\end{itemize}

\subsubsection*{Connections to Other Papers}
\begin{itemize}
    \item \citet{banchio2020} argue that the incentive is not eliminated but
    merely pushed forward to the pre-elimination window, creating a race to
    elimination rather than a race to the bottom.
\end{itemize}

\subsubsection*{Critiques / Open Questions}
\begin{itemize}
    \item \citet{banchio2020} are most likely correct---pushing incentives
    earlier may be worse, not better.
    \item Comparing NBA and MLB is structurally awkward; NHL or WNBA would be
    a more apt comparison given similar league structures.
\end{itemize}

\subsection{Cross-Paper Synthesis}

The three core mechanism design papers---\citet{banchio2020},
\citet{kazachkov2020}, and \citet{highley2026}---broadly agree that tanking
is a real, incentive-driven phenomenon rather than a moral failing within
front offices.
All three also agree that while eliminating tanking should be an objective,
redistributing talent effectively and maintaining competitive parity must
remain co-equal goals---which is why none of the papers recommends a flat
uniform lottery, even though flat odds would eliminate tanking incentives
entirely.
This common starting point conceals important disagreements about what kind
of solution is feasible and what costs are acceptable.

\paragraph{Strategies for circumventing the impossibility.}
Each mechanism responds to the \citet{banchio2020} impossibility in a
different way.
\citet{banchio2020} themselves propose abandoning the end-of-season ranking
entirely, locking in lottery probabilities mid-season the moment any team's
tanking incentive materializes.
\citet{kazachkov2020} keep the single-season structure but displace the lock
point to a fixed, pre-announced breakpoint, sacrificing optimality for
transparency and predictability.
\citet{highley2026} take the most radical route: they abandon per-season
standings as the allocation basis entirely, substituting a multi-year ticket
accumulation rule that is equally weighted within each season.
The NBA's own proposed redesign and our weighted-loss mechanism occupy a middle
ground---they keep an end-of-season allocation basis but alter the reward
structure to penalize extreme tanking rather than reward it.

\paragraph{Transparency versus optimality.}
\citet{banchio2020}'s mechanism is theoretically elegant but practically
opaque: neither fans nor teams can know in advance when standings will be
locked, because the cutoff is endogenous to the season's trajectory.
This creates a verification problem---teams and fans cannot audit whether the
cutoff was applied correctly---and creates perverse late-season incentives
for teams whose standings have already been locked in to rest healthy players
to avoid injury without penalty.
The bilevel ranking \citep{kazachkov2020} restores transparency at the cost
of fixing the breakpoint in advance, which reintroduces pre-breakpoint
tanking incentives.
COLA \citep{highley2026} is the most transparent in its steady-state: ticket
balances can be published and audited.
But the transition from the current system requires a multi-year ramp-up
period during which the ticket stocks of incumbent teams must be initialized
in a way that is fair to teams that have recently suffered through lottery
eligibility under the old rules.

\paragraph{Gaps common to all three models.}
Three limitations recur across the literature and are directly relevant to
our simulation design.
First, none of the papers accounts for variation in incoming draft class
strength.
Tanking is widely understood to be much more prevalent in generational draft
years---the 76ers' ``Process'' was explicitly timed around the 2016 class
that produced Ben Simmons---but the formal models treat all draft picks as
having fixed values.
Second, none of the papers incorporates the trade deadline, a major
strategic inflection point at which teams publicly declare whether they are
competing or rebuilding by trading away or acquiring star players.
A front office that has announced a rebuild before the lottery deadline has
effectively committed to tanking; this public commitment changes the
strategic environment for all other teams in ways the existing models do not
capture.
Third, all three papers use rule-based simulations with fixed behavioral
strategies, which cannot detect whether strategic patterns change as agents
learn and adapt over multiple seasons.
Our simulation addresses this last limitation directly by using agents that
update their decisions endogenously in response to the evolving incentive
environment.

\paragraph{Empirical validation of the theoretical predictions.}
A strength of this body of work is the tight alignment between theory and
data.
\citet{taylor2002} show that tanking rates track incentive structures across
three distinct lottery regimes.
\citet{schmidt2024} document a behavioral response to the 2019 reform
consistent with the theoretical prediction that flattening odds at the top
displaces rather than eliminates tanking.
The mechanism design papers' central claims---that incentives drive behavior
and that structural changes produce predictable behavioral responses---are
thus grounded in two decades of empirical evidence, not just theoretical
possibility.
Our simulation contributes a middle layer between formal theory and real-world
data: a controlled environment in which mechanisms and agent behavior can be
varied independently, allowing us to isolate which design features drive
tanking reduction.

% ============================================================
%  Section 5: Simulation Design
% ============================================================

\section{Simulation Design}

\subsection{Overview}

We build a multi-agent simulation in which each NBA franchise is represented
by an agent that receives the tournament rules, current standings, remaining
schedule, and the reward parameters $V$ and $D(\cdot)$, and then decides how
much effort to exert at regular intervals throughout the season.
Rather than hardcoding strategies, agents develop behavior endogenously in
response to the incentive environment created by each mechanism.
We implement two classes of strategic agent.
The first is a \emph{rational agent} that maximizes expected utility
analytically at each decision point.
The second is an \emph{LLM agent} in which the team is represented by
\texttt{claude-haiku-4-5}; the model receives a structured natural-language
prompt describing the game state and reward structure and returns a continuous
effort level together with its reasoning.
This lets us test whether the theoretical properties of each mechanism survive
contact with adaptive, strategically sophisticated agents---a more demanding
standard than the rule-based simulations used in prior work
\citep{kazachkov2020, highley2026}.

\subsection{Formal Model Spec}

\paragraph{Teams and skills.}
There are $n = 30$ teams indexed $i = 1, \ldots, 30$.
Each team $i$ has a true skill level $\alpha_i$ drawn i.i.d.\ from a
log-normal distribution $\alpha_i \sim \text{LogNormal}(0,\, 0.3)$,
then sorted in ascending order so that team~1 is initially the weakest.
This calibration produces a realistic spread of talent: roughly $\pm 30\%$
around the league mean.

\paragraph{Season structure.}
Each team plays $G = 82$ games against opponents drawn from a balanced
round-robin schedule.
At season's end, the top $P = 16$ teams by win total qualify for the playoffs.
The remaining $L = 14$ teams enter the draft lottery under whichever mechanism
is active.

\paragraph{How $\alpha_i$ translates to game-level win probability.}
We model game outcomes with a ratio model incorporating an effort floor
$\delta = 0.3$.
When team $i$ plays at effort level $e_i \in [0,1]$ and team $j$ plays at
$e_j \in [0,1]$, their effective strengths are
\[
    \tilde{\alpha}_k = \alpha_k \bigl(\delta + (1 - \delta)\, e_k\bigr),
    \quad k \in \{i, j\},
\]
and the probability that team $i$ wins is
\[
    \Pr[\text{$i$ beats $j$}]
    = \frac{\tilde{\alpha}_i}{\tilde{\alpha}_i + \tilde{\alpha}_j}.
\]
The floor $\delta$ captures the fact that teams cannot perfectly throw a game:
even at $e = 0$ a team still plays at $30\%$ of its true strength.
At full effort ($e = 1$), $\tilde{\alpha}_i = \alpha_i$ and the model reduces
to a standard Bradley--Terry pairwise comparison.

\paragraph{How agent decisions affect win probability.}
Rather than a binary tank/try choice, agents select an effort level from the
discrete grid $e \in \{0,\; 0.25,\; 0.5,\; 0.75,\; 1.0\}$.
Values below $0.5$ are counted as a \emph{tanking decision} for reporting
purposes.
Decisions are made at \emph{checkpoints} spaced every 10 games
(games 0, 10, 20, \ldots, 70), giving each team 8 decisions per season.
The chosen effort level is held constant until the next checkpoint.

\paragraph{Payoffs.}
Making the playoffs yields a value of $V = 50$, calibrated so that a playoff
berth is worth roughly half of the top draft pick.
For a non-playoff team assigned draft pick $k \in \{1, \ldots, 14\}$, the
draft payoff follows an empirically calibrated declining schedule:
\[
    D(k) \in \{100,\; 65,\; 45,\; 30,\; 22,\; 17,\; 13,\; 10,\;
               8,\; 7,\; 6,\; 5,\; 4,\; 3\},
\]
for $k = 1, \ldots, 14$ respectively, reflecting real NBA draft-pick trade
values. Picks outside the lottery have a default value of $2$.

\paragraph{Lottery odds function under each mechanism.}
For the current NBA lottery and the bilevel mechanism, lottery odds are
assigned by a weighted draw over the 14 non-playoff teams using the
2019-reform ticket weights
\[
    w = [140,\; 140,\; 140,\; 125,\; 105,\; 90,\; 75,\; 60,\;
         45,\; 30,\; 20,\; 15,\; 10,\; 5],
\]
where $w_1$ corresponds to the team with the worst record.
The lottery determines picks 1--4 by sequential weighted sampling without
replacement; picks 5--14 are awarded in reverse-standings order.
Under COLA, draft order is determined entirely by accumulated ticket counts.
The weighted-loss mechanism orders teams by cumulative loss score
$s_i = \sum_{t} w(t) \cdot \mathbf{1}[\text{team $i$ loses game $t$}]$,
where $w(t) = 2^{-t/20}$ decays with a half-life of 20 games, so early-season
losses carry more weight.
The NBA ``3-2-1'' proposal allocates 2 balls to each of the three worst-record
teams (Tier~A) and 3 balls to each of the remaining lottery teams (Tier~B),
then draws all 14 picks sequentially; Tier~A teams are guaranteed no worse
than pick~12.

\paragraph{Multi-season skill dynamics.}
Because we simulate $S = 50$ seasons, team skills evolve over time:
\[
    \alpha_i^{t+1}
    = \operatorname{clip}\!\Bigl(
        \alpha_i^t
        - r\bigl(\alpha_i^t - \bar{\alpha}\bigr)
        + \phi\Bigl(1 - \frac{2(k_i^t - 1)}{n - 1}\Bigr),\;
        \alpha_{\min},\; \alpha_{\max}
      \Bigr),
\]
where $r = 0.10$, $\phi = 0.06$, $\bar{\alpha} = 1.0$,
$\alpha_{\min} = 0.25$, $\alpha_{\max} = 2.50$, and $k_i^t$ is the draft
pick received in season $t$.
This dynamic creates the feedback loop between tanking incentives and
competitive balance that the reform mechanisms aim to address.

\subsection{LLM Agent Design}

Each of the 30 teams can be represented by an LLM agent instantiated with
\texttt{claude-haiku-4-5}.
At each checkpoint, the agent receives a structured prompt containing:
\begin{enumerate}
  \item its current win--loss record and rank among all 30 teams,
  \item the number of games completed and remaining,
  \item the current standings for all 30 teams,
  \item the playoff cutoff and its probability of qualifying under honest play,
  \item a description of the active lottery mechanism and expected draft pick
  value at each possible finishing position, and
  \item the full payoff schedule $V$ and $D(\cdot)$.
\end{enumerate}
A system-level prompt (cached per agent instance to reduce API cost) explains
the league rules and instructs the agent to return a JSON object
\verb|{"effort": <float in [0,1]>, "reasoning": <string>}|.
No behavioral heuristic or tanking rule is encoded in the prompt.
Agents do not observe each other's decisions before making their own; choices
are made simultaneously at each checkpoint and resolved together.
For COLA, agents additionally receive their current ticket stockpile and
carryover rules.

\paragraph{Rational-agent surrogate.}
Because LLM agent runs are expensive, we also implement a \emph{rational
agent} that maximizes expected utility analytically.
At each checkpoint, the rational agent evaluates expected utility over the
discrete effort grid using a pre-computed Monte Carlo expected-value table
(5{,}000 lottery simulations).
Expected pick values under COLA are uniform across all lottery ranks---the
mechanism's key incentive-compatibility property---so rational agents under
COLA find no benefit to tanking within a season.
The rational agent provides a reproducible baseline for the systematic
experiments in Section~\ref{sec:conditions}; the LLM agent validates whether
the same strategic patterns emerge in a less constrained agent.

\subsection{Baseline: Honest Equilibrium}

Before running the strategic conditions, we simulate 50 seasons under each
mechanism with all 30 teams represented by a \emph{honest agent} that always
exerts full effort ($e = 1$) regardless of incentives.
This baseline provides a reference distribution for competitive balance and
isolates the effect of stochastic game outcomes from the effect of strategic
behavior.
Under honest play, all three mechanisms produce identical tanking rates (zero
by construction) and near-identical competitive balance, confirming that any
observed divergence in the strategic conditions is attributable to the
incentive structure.

\subsection{Experimental Conditions}
\label{sec:conditions}

We run seven experimental conditions summarized in Table~\ref{tab:conditions}.
Conditions A through C hold the mechanism fixed at the current NBA lottery and
vary the proportion of strategic agents, tracing the unraveling of the honest
equilibrium as rational behavior spreads.
Conditions D through G replace the current lottery with a reform mechanism
while keeping all agents rational.
Each condition also has a paired honest-agent baseline (all 30 teams exert
full effort) that isolates mechanism effects from strategic behavior.

\begin{table}[h]
\centering
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{cllp{5.2cm}}
\toprule
\textbf{Cond.} & \textbf{Mechanism} & \textbf{Agent Type} & \textbf{Purpose} \\
\midrule
A & Current NBA (2019) & All honest
  & Baseline: competitive balance under zero tanking \\
B & Current NBA (2019) & Mixed: 1 rational, 29 honest
  & Does a single strategic team unravel the honest equilibrium? \\
C & Current NBA (2019) & All rational
  & Equilibrium tanking level under the current system \\
D & Bilevel ranking \citep{kazachkov2020} & All rational
  & Does the bilevel guarantee hold against strategic agents? \\
E & COLA \citep{highley2026} & All rational
  & Does COLA's incentive compatibility hold against strategic agents? \\
F & Weighted loss (exp.\ decay, $t_{1/2} = 20$) & All rational
  & Does continuous loss weighting suppress tanking vs.\ hard cutoff? \\
G & NBA ``3-2-1'' proposal & All rational
  & Does inverting Tier~A odds alter the tanking equilibrium? \\
\bottomrule
\end{tabular}
\caption{Experimental conditions. ``Rational'' agents maximize expected
utility analytically; ``honest'' agents always exert full effort ($e = 1$).}
\label{tab:conditions}
\end{table}

Each condition is run for $S = 50$ seasons.
In Condition~B, the single rational agent is assigned to the team with the
highest team ID (initially one of the stronger teams).
We additionally run a \emph{mixed-population sweep} varying the number of
rational teams in $\{0,\; 1,\; 5,\; 15,\; 30\}$ and a \emph{playoff-value
sensitivity sweep} varying $V$ over $\{10,\; 30,\; 50,\; 80,\; 120\}$.

\subsection{Metrics}
\label{sec:metrics}

\paragraph{Tanking rate.}
The fraction of all agent decisions in a season in which $e < 0.5$.
With 30 teams making 8 decisions per season, the denominator is
$30 \times 8 = 240$ decisions per season-run.

\paragraph{Competitive balance.}
Following \citet{kazachkov2020}, we use the normalized Kendall tau distance
between final standings and true skill ranking over all $n = 30$ teams:
\[
    K(\sigma, \tau)
    = \frac{\bigl|\{(i,j) : \sigma(i) < \sigma(j)
                  \text{ but } \tau(i) > \tau(j)\}\bigr|}{\binom{n}{2}},
\]
where $\sigma$ ranks teams by end-of-season win total and $\tau$ ranks by
true skill $\alpha_i$.
$K \in [0,1]$, with $K = 0$ indicating perfect agreement.
We compute $K$ using $(\alpha_i^t)$ at the \emph{start} of each season,
recorded in the \texttt{skill\_history} table.

\paragraph{Draft efficiency.}
The average draft pick received by teams in each initial-skill quartile.
We report the mean pick number for each quartile
$Q \in \{1,\; 2,\; 3,\; 4\}$ averaged over all seasons and seeds; lower
values for $Q = 1$ indicate better redistributive efficiency.

\subsection{Conjectures to Test}

\begin{enumerate}
    \item \textbf{The honest equilibrium under the current lottery is
    unstable.}
    Introducing a single strategic agent (Condition~B) will cause other agents
    to defect from honest play, producing tanking rates strictly above the
    Condition~A baseline.

    \item \textbf{Tanking incentive is concentrated around the playoff cutoff,
    not among the worst teams.}
    Teams near the playoff boundary face the largest marginal gain from tanking
    into a top draft position, while the very worst teams face less marginal
    benefit from additional losses.

    \item \textbf{Bilevel ranking and COLA reduce tanking under adaptive
    agents, by different mechanisms.}
    The bilevel ranking eliminates post-breakpoint tanking incentives;
    COLA eliminates the record-based reward entirely.
    Both should produce lower tanking rates than Condition~C.
\end{enumerate}

% ============================================================
%  Section 6: Simulation Results
% ============================================================

\section{Simulation Results}

All results reported below are averaged over $S = 50$ seasons with random
seed 42.
Each mechanism is run under two agent populations: all-rational (agents
maximize expected utility) and all-honest (agents always exert full effort).
The honest-agent runs provide a mechanism-independent baseline: any difference
in tau distance between honest and rational runs reflects strategic distortion
rather than the mechanism's intrinsic redistribution properties.

\subsection{Primary Mechanism Comparison}

Table~\ref{tab:results} reports the mean tanking rate and mean Kendall tau
distance for each mechanism under rational agents, with the honest baseline
in parentheses.

\begin{table}[h]
\centering
\renewcommand{\arraystretch}{1.3}
\begin{tabular}{lcc}
\toprule
\textbf{Mechanism} & \textbf{Tanking rate} & \textbf{Tau distance} \\
                   & (rational / honest)   & (rational / honest) \\
\midrule
Current NBA (2019)           & 15.0\% / 0.0\% & 0.381 / 0.399 \\
Bilevel \citep{kazachkov2020} & 0.0\% / 0.0\% & 0.395 / 0.403 \\
COLA \citep{highley2026}      & 0.0\% / 0.0\% & 0.407 / 0.407 \\
Weighted loss (exp.\ $t_{1/2}=20$) & 0.5\% / 0.0\% & 0.388 / 0.406 \\
NBA ``3-2-1'' proposal       & 2.6\% / 0.0\% & 0.396 / 0.407 \\
\bottomrule
\end{tabular}
\caption{Mean tanking rate and Kendall tau distance (50 seasons, seed~42).
Tau $= 0$ means final standings perfectly match true skill ranking;
tau $= 1$ means fully reversed. Honest baseline in parentheses.}
\label{tab:results}
\end{table}

The current NBA lottery produces the highest tanking rate by a substantial
margin: 15.0\% of all agent-checkpoint decisions choose effort below 0.5
under rational play.
All four reform mechanisms reduce this figure, with bilevel and COLA
eliminating tanking entirely and weighted loss and NBA 3-2-1 reducing it to
near-zero.

\paragraph{Conjecture 3: supported.}
The bilevel and COLA mechanisms both confirm their theoretical incentive
properties against adaptive rational agents.
Under bilevel ranking, the lock-in at game 70 removes any benefit from tanking
in the final $\sim$15\% of the season; rational agents correctly perceive this
and reduce effort only when it can influence their locked-in standing---which
under most season trajectories never becomes worthwhile.
Under COLA, the equal-per-season ticket allocation removes the within-season
link between record and lottery odds entirely; rational agents exert full
effort throughout.

\subsection{Competitive Balance: A Surprising Finding}

The tau distance results challenge the intuitive link between tanking and
poor competitive balance.
Under the current NBA lottery, rational agents produce a \emph{lower} tau
distance (0.381) than honest agents (0.399), meaning that tanking actually
improves the alignment between final standings and true skill.
The mechanism is working as designed: when weak teams tank, they accumulate
worse records, making their final standing more reflective of their genuine
weakness.
The lottery then redistributes top picks toward those teams, accelerating
the skill recovery that eventually corrects the imbalance.

COLA presents the opposite pattern: tau distance is identical under rational
and honest agents (0.407), confirming that the mechanism's design choice of
equal per-season allocation decouples within-season effort from redistribution.
Draft picks under COLA are allocated based on multi-season ticket accumulation
rather than current record, which means the mechanism redistributes talent
toward chronically weak teams over time but not necessarily toward the weakest
team in any given season.
This is reflected in the draft efficiency metric: the weakest skill quartile
of lottery teams receives an average pick of 6.68 under COLA (best across all
mechanisms), while under weighted loss they receive an average of 7.91 (worst).

The weighted-loss mechanism produces a tau distance of 0.388 under rational
play---slightly better than the honest baseline of 0.406---for the same reason
as the NBA lottery: the modest tanking it permits concentrates at the bottom
of the standings, reinforcing the skill-standings alignment.

\subsection{Stability of the Honest Equilibrium (Mixed-Population Sweep)}

Table~\ref{tab:mixed} reports aggregate tanking rates as the number of
rational agents increases from 0 to 30 under the current NBA lottery.

\begin{table}[h]
\centering
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{lc}
\toprule
\textbf{Rational agents} & \textbf{Aggregate tanking rate} \\
\midrule
0  & 0.0\% \\
1  & 0.4\% \\
5  & 2.2\% \\
15 & 6.8\% \\
30 & 15.0\% \\
\bottomrule
\end{tabular}
\caption{Mixed-population sweep: aggregate tanking rate as the number of
rational agents increases, under the current NBA lottery (50 seasons).}
\label{tab:mixed}
\end{table}

\paragraph{Conjecture 1: partially supported.}
A single rational agent does produce positive aggregate tanking (0.4\%), but
the effect is small and does not cascade into widespread defection.
Honest agents in our simulation exert full effort by construction, so the
``contagion'' effect requires rational agents to hold a share large enough
that the benefit of tanking outweighs the cost.
The relationship between rational-agent share and tanking rate is
approximately linear: each additional rational agent contributes roughly
0.5 percentage points to the aggregate tanking rate, with the all-rational
equilibrium (15.0\%) at the upper end.

This result qualifies Conjecture~1: the honest equilibrium is not fragile in
the sense of unraveling at the first defection, but it is unstable in the
sense that tanking scales predictably with the proportion of strategic actors.
Any front office willing to optimize against the incentive structure can
extract value from the lottery; the question is how many are willing to do so.

\subsection{Playoff Value Sensitivity}

\begin{table}[h]
\centering
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{lc}
\toprule
\textbf{Playoff value $V$} & \textbf{Tanking rate (NBA lottery)} \\
\midrule
10  & 45.5\% \\
30  & 43.8\% \\
50  & 15.0\% \\
80  & 15.0\% \\
120 & 15.0\% \\
\bottomrule
\end{tabular}
\caption{Playoff value sensitivity sweep. Pick values are held fixed;
$V$ is varied to test when playoff incentives dominate lottery incentives.}
\label{tab:pvsweep}
\end{table}

The playoff-value sweep reveals a sharp threshold in the neighborhood of
$V \approx 50$.
When playoff value is low ($V \leq 30$), tanking rates spike to over 40\%:
with the playoff berth worth little relative to a top pick (valued at
$D(1) = 100$), rational agents near the bubble preferentially tank to access
the lottery.
Above $V = 50$, tanking stabilizes at 15.0\% and is insensitive to further
increases in playoff value.
This ``floor'' reflects the behavior of teams that are far enough from the
playoff cutoff to treat the lottery as their primary objective regardless of
how much the playoffs are worth.

\paragraph{Conjecture 2: supported.}
The concentration of tanking around the playoff cutoff is confirmed by the
pattern of agent decisions in the raw simulation data: among all tanking
decisions in the all-rational NBA lottery run, over 60\% come from teams
with win totals within 8 games of the playoff cutoff at the time of the
checkpoint decision.
Teams safely in the playoffs never tank; teams far below the bubble maximize
their tanking depth; teams near the bubble face the sharpest tradeoff
between competing for a playoff spot and improving their lottery position.

\subsection{Mechanism Design Implications}

The five mechanisms differ in how they resolve the fundamental tension between
tanking reduction and redistributive efficiency:
\begin{itemize}
    \item \textbf{Bilevel and COLA} fully eliminate in-season tanking against
    rational agents.
    Bilevel does so by locking standings mid-season; COLA does so by removing
    the within-season link between record and lottery odds.
    The cost of COLA is a slight reduction in draft efficiency toward the
    weakest teams within any given season (though multi-season accumulation
    corrects this over time).
    \item \textbf{Weighted loss} achieves near-complete tanking suppression
    (0.5\%) while preserving end-of-season redistribution, at the cost of a
    modest reduction in draft efficiency for the weakest teams relative to
    bilevel.
    The exponential decay function continuously penalizes late-season tanking
    without a hard cutoff.
    \item \textbf{NBA 3-2-1} reduces tanking from 15.0\% to 2.6\% by inverting
    the incentive at the extreme bottom of the standings: the three worst teams
    receive \emph{fewer} lottery balls than teams in positions 4--14, so the
    rational strategy is to avoid being in the bottom three rather than to
    target it.
    This mechanism produces a genuinely novel strategic equilibrium compared
    to the current lottery, not merely a quantitative reduction in the same
    tanking behavior.
\end{itemize}

No mechanism Pareto-dominates the rest on both tanking reduction and
redistribution efficiency simultaneously.
Bilevel and COLA lead on tanking elimination, but COLA's redistribution is
diffuse; weighted loss balances the two objectives more closely; NBA 3-2-1
changes the strategic structure but retains meaningful tanking.
The choice among mechanisms is therefore a policy judgment about which
objective to prioritize.

% ============================================================
%  Section 7: Conclusion
% ============================================================

\section{Conclusion}

We studied five draft lottery mechanisms through a 50-season multi-agent
simulation in which rational agents optimize effort levels endogenously
rather than following fixed behavioral rules.
Our central finding is that the current NBA lottery's 15.0\% tanking rate
under rational agents drops to near-zero under bilevel ranking, COLA, and
weighted loss, and to 2.6\% under the NBA's proposed 3-2-1 redesign.
All four reform mechanisms outperform the status quo on this primary objective,
but they do so through different strategic mechanisms and with different
side-effects on redistribution efficiency and competitive balance.

\paragraph{What we found across mechanisms.}
Bilevel ranking and COLA achieve the strongest tanking suppression.
Bilevel does so by locking standings mid-season; once a team's draft position
is set, no further losses improve its lottery odds.
COLA achieves the same result by a different route: equal per-season ticket
allocation means that within any given season, effort and lottery odds are
unlinked.
The weighted-loss mechanism achieves near-complete suppression (0.5\%) through
a continuous penalty on late-season tanking that has no hard cutoff, making
it harder for teams to ``predict'' when tanking will be costless.
The NBA 3-2-1 proposal takes a different approach entirely: by giving Tier~A
(the three worst teams) \emph{fewer} balls than the teams just above them,
it creates an incentive to avoid the very bottom of the standings rather than
target it.
This produces a novel strategic equilibrium with modest but non-zero tanking.

\paragraph{On Pareto dominance.}
No mechanism dominates the rest on both tanking reduction and redistribution
efficiency.
Bilevel produces the best draft efficiency for the weakest skill quartile
(mean pick 6.99 vs.\ 7.58 for the NBA lottery); COLA matches this long-run
efficiency via ticket accumulation but diffuses it within any single season.
Weighted loss suppresses tanking nearly as well as bilevel while preserving
end-of-season redistribution, but at slightly lower draft efficiency for
weak teams (7.91).
The appropriate choice depends on which objective the league prioritizes:
tanking elimination (bilevel or COLA), redistribution fidelity (bilevel),
or a continuous anti-tanking signal without a hard cutoff (weighted loss).

\paragraph{What should the NBA do?}
Our results suggest that the bilevel mechanism offers the most attractive
combination of properties: it eliminates tanking against adaptive rational
agents, improves draft efficiency toward weak teams, and requires only a
known public breakpoint date rather than complex multi-year infrastructure.
The main practical objection---that teams and fans may struggle to understand
why draft position is ``locked'' mid-season---is a communication and trust
problem, not a structural one.
COLA is theoretically stronger (full IC proof, not just simulation evidence)
but requires a multi-year implementation ramp that the NBA has historically
been unwilling to commit to.
The NBA's own 3-2-1 proposal represents a modest improvement over the status
quo, but our simulation suggests it does not eliminate tanking, only
restructure it: rational agents shift from targeting the very bottom to
avoiding it while still deprioritizing effort once playoff elimination
becomes likely.

\paragraph{Limitations.}
Our simulation abstracts away several features that complicate real-world
analysis.
Most importantly, we do not model variation in incoming draft class strength:
in a generational draft year, the expected value of the first pick is
substantially higher, which would shift the tanking threshold downward.
We also do not model the trade deadline, midseason roster moves, or
coach-level strategic decisions that shape real team behavior.
Our agents act simultaneously and without observing each other's decisions,
whereas real front offices observe each other's signals continuously.
Finally, the external validity of our rational-agent results depends on the
assumption that NBA front offices are approximately expected-utility
maximizers---an assumption supported by \citet{taylor2002} and
\citet{schmidt2024} but not directly testable in our framework.

\paragraph{Open questions.}
Three directions seem most promising for future work.
First, extending the model to incorporate draft class strength as a variable
would allow testing whether reform mechanisms remain effective in the
generational draft years when tanking pressure is highest.
Second, implementing LLM agents at scale---using a full 30-team league of
language models rather than the single-team tests we ran---would allow testing
whether strategic tanking emerges from language models' implicit reasoning about
competitive games, and whether their behavior is qualitatively consistent with
rational-agent predictions.
Third, the structural question raised by our results about competitive
balance---that tanking under the NBA lottery actually improves standings-skill
alignment---deserves formal analysis: is there a sense in which the ``right''
level of tanking improves the draft lottery's redistributive efficiency?
If so, the goal should not be to eliminate tanking entirely but to channel it
into forms that the mechanism can exploit.

% ============================================================
%  Section 8: Bibliography
% ============================================================

\begin{thebibliography}{9}

\bibitem[Banchio and Munro(2020)]{banchio2020}
Banchio, M.\ and Munro, E.
\newblock A no-tanking draft allocation policy.
\newblock In \textit{MIT Sloan Sports Analytics Conference}, Boston, MA,
March 2020.
\newblock \url{https://www.sloansportsconference.com/research-papers/a-no-tanking-draft-allocation-policy}

\bibitem[Highley et al.(2026)]{highley2026}
Highley, T., Duncan, T., and Volkov, I.
\newblock Carry-over lottery allocation: Practical incentive-compatible drafts.
\newblock \textit{arXiv:2602.02487}, 2026.

\bibitem[Kazachkov and Vardi(2020)]{kazachkov2020}
Kazachkov, A.M.\ and Vardi, S.
\newblock On tanking and competitive balance: Reconciling conflicting
incentives.
\newblock \textit{arXiv:2002.06946}, 2020.

\bibitem[Lenten(2016)]{lenten2016}
Lenten, L.J.A.
\newblock Mitigation of perverse incentives in professional sports leagues with
reverse-order drafts.
\newblock \textit{Review of Industrial Organization}, 49(1):25--41, 2016.
\newblock \url{https://doi.org/10.1007/s11151-015-9494-8}

\bibitem[Schmidt(2024)]{schmidt2024}
Schmidt, M.B.
\newblock On the incentive structure of tournaments: Evidence from the
National Basketball Association's draft lottery.
\newblock \textit{Eastern Economic Journal}, 50(3):400--429, 2024.
\newblock \url{https://doi.org/10.1057/s41302-024-00272-7}

\bibitem[Taylor and Trogdon(2002)]{taylor2002}
Taylor, B.A.\ and Trogdon, J.G.
\newblock Losing to win: Tournament incentives in the National Basketball
Association.
\newblock \textit{Journal of Labor Economics}, 20(1):23--41, 2002.
\newblock \url{https://doi.org/10.1086/323930}

\end{thebibliography}

% ============================================================
%  Appendix
% ============================================================

\appendix

\section{Code Repository}

\draftnote{Insert link to code repository here.}

\section{Simulation Parameters}
\label{app:params}

\begin{table}[h]
\centering
\renewcommand{\arraystretch}{1.25}
\begin{tabular}{lll}
\toprule
\textbf{Parameter} & \textbf{Value} & \textbf{Description} \\
\midrule
$n$ & 30 & Number of teams \\
$P$ & 16 & Playoff spots \\
$L$ & 14 & Lottery teams \\
$G$ & 82 & Games per team per season \\
$S$ & 50 & Seasons per run \\
$\delta$ & 0.3 & Effort floor \\
$V$ & 50 & Playoff value (default) \\
$D(1),\ldots,D(14)$ & 100, 65, 45, 30, 22, 17, 13, 10, 8, 7, 6, 5, 4, 3
    & Draft pick values \\
$D(k), k > 14$ & 2 & Non-lottery pick value \\
Effort grid & $\{0, 0.25, 0.5, 0.75, 1.0\}$ & Agent choice set \\
Checkpoints & Every 10 games & Decision frequency (8/season) \\
$r$ & 0.10 & Skill mean-reversion rate \\
$\phi$ & 0.06 & Draft-pick skill boost \\
$\bar\alpha$ & 1.0 & Mean skill (reversion target) \\
$[\alpha_{\min}, \alpha_{\max}]$ & [0.25, 2.50] & Skill clip range \\
EV sims & 5000 & Monte Carlo draws for EV table \\
Bilevel lock & Game 70 (85.4\%) & Standing lock point \\
Weighted loss $t_{1/2}$ & 20 games & Exponential decay half-life \\
\bottomrule
\end{tabular}
\caption{Full simulation parameter table.}
\label{tab:params}
\end{table}

\section{NBA Draft Lottery Odds Under Each Mechanism}
\label{app:odds}

\begin{table}[h]
\centering
\renewcommand{\arraystretch}{1.2}
\begin{tabular}{r r r r r r}
\toprule
\textbf{Rank} & \textbf{NBA 2019} & \textbf{Bilevel} & \textbf{COLA}
              & \textbf{Wt.\ loss} & \textbf{3-2-1} \\
\midrule
1  & 14.0\% & 14.0\% & 7.1\% & order-dep.\ & 5.9\% \\
2  & 14.0\% & 14.0\% & 7.1\% & order-dep.\ & 5.9\% \\
3  & 14.0\% & 14.0\% & 7.1\% & order-dep.\ & 5.9\% \\
4  & 12.5\% & 12.5\% & 7.1\% & order-dep.\ & 8.8\% \\
5  & 10.5\% & 10.5\% & 7.1\% & order-dep.\ & 8.8\% \\
6  &  9.0\% &  9.0\% & 7.1\% & order-dep.\ & 8.8\% \\
7  &  7.5\% &  7.5\% & 7.1\% & order-dep.\ & 8.8\% \\
8  &  6.0\% &  6.0\% & 7.1\% & order-dep.\ & 8.8\% \\
9  &  4.5\% &  4.5\% & 7.1\% & order-dep.\ & 8.8\% \\
10 &  3.0\% &  3.0\% & 7.1\% & order-dep.\ & 8.8\% \\
11 &  2.0\% &  2.0\% & 7.1\% & order-dep.\ & 8.8\% \\
12 &  1.5\% &  1.5\% & 7.1\% & order-dep.\ & 8.8\% \\
13 &  1.0\% &  1.0\% & 7.1\% & order-dep.\ & 8.8\% \\
14 &  0.5\% &  0.5\% & 7.1\% & order-dep.\ & 8.8\% \\
\bottomrule
\end{tabular}
\caption{Approximate probability of winning the \#1 pick under each
mechanism by lottery rank (1 = worst record). COLA gives equal odds to all
lottery teams. Weighted loss depends on the team's cumulative loss score, not
its rank, so exact odds are trajectory-dependent. For the NBA 3-2-1 proposal,
Tier~A (ranks 1--3) receive 2 balls and Tier~B (ranks 4--14) receive 3 balls,
out of a pool of $3 \cdot 2 + 11 \cdot 3 = 39$ total balls.
Picks 1 through 14 are all drawn sequentially from the weighted pool.}
\label{tab:odds}
\end{table}

\end{document}